---
title: 'Statistics Case Study: Cross Fell Grazing 2013: Analysis'
author: "CK"
date: "7 Sept 2017"
output: html_notebook
---

###Time run
Time run: `r Sys.time()`

###Packages
```{r}
library(tidyverse)
```


###Load data

```{r}
load(file="CrossFell.data.prepd.rda")
```


#Replicate report

##4.2	Vegetation condition 

###4.2.1   Dominant Graminoids

> In 2013, sheep’s fescue (Festuca ovina) was recorded as being the dominant graminoid in 14 (25%) of quadrats (figure 2), with an average sward height of 2.86 cm (table 1). Stiff sedge (Carex bigelowii) was recorded as dominant in 33(59%) quadrats, with an average sward height of 6.94 cm.  From figure 2 it can be seen that the proportion of quadrats in which sheep’s fescue is dominant has steadily decreased from 81% in 2003, whilst the proportion in which stiff sedge is judged to be dominant has increased from 14.3%.

> Figure 2.  The proportion of quadrats in which sheep’s fescue and stiff sedge are dominant in each survey, 2003 – 2010.

Cannot replicate this, because we don't have the species data available.   
 
> Table 1.  Mean height of key species groups measured in each survey year

Species |2003	|2005	|2008	|2010	|2013
--- |--- |--- |--- |--- |---
Bilberry	|2.4a	|2.88a	|2.58a	|4.66c |3.53b
Sheep’s fescue	|3.11a	|3.78b	|3.88ab |3.89ab |2.86a
Stiff sedge	|4.2a	|7.13b	|6.56b	|7.56b	|6.94b
Moss/ lichen	|1.95a	|2.37a |3.48b |3.48b |3.56b
All graminoids	|3.28a |4.46b	|5.04bc	|6.45d |5.75cd

Below is the same table, recalculated in R but missing Stiff sedge as I don't have the data for that.  I don't know what the letters mean: 
```{r, echo=FALSE}
Table1 <- spread(data = select(.data = ht.means, taxon, year, mean), 
                 key = year, 
                 value = mean) 
names(Table1)[1] <- "Species"
Table1 <- Table1 %>% 
  as.data.frame() %>% 
  mutate(Species = factor(c("Sheep's Fescue", "All graminoids", "Moss/ lichen", "Bilberry"), 
                          levels = c("Bilberry", "Sheep's Fescue", "Moss/ lichen", "All graminoids"))) %>% 
  arrange(Species)

knitr::kable(Table1, digits = 2, caption = "Table 1: Mean height of key species groups measured in each survey year")
```


####Carry out ANOVA on just one taxon

```{r}
aov.moss <- aov(data = dplyr::filter(ht.data, taxon == "moss"), formula = height ~ year)
anova(aov.moss)
print(model.tables(aov.moss, "means"), digits = 3)
model.tables(x = aov.moss)
tuk.moss <- TukeyHSD(aov.moss)
tuk.moss
plot(tuk.moss)

tuk.moss.df <- data.frame(tuk.moss$year) %>%
  rownames_to_column("year") %>% 
  separate(col = year, into = c("year2", "year1"), sep = "-") %>% 
  mutate(year.diff = as.numeric(year2) - as.numeric(year1)) %>% 
  arrange(p.adj) %>% 
  filter(p.adj<0.05)
tuk.moss.df  


```

####Carry out ANOVA on just one taxon using pckg::userfriendlyscience

Not evaluated as it doesn't work except on R v.3.4.0

```{r, eval=FALSE}
library(userfriendlyscience)

#oneway(data = dplyr::filter(ht.data, taxon == "moss"), formula = height ~ year)
ow.moss <- oneway(y = ht.data$height, x = ht.data$year, posthoc = "tukey", means = TRUE, fullDescribe = TRUE, plot = TRUE, pvalueDigits = 3)
#class(ow.moss)
ow.moss$output

ow.moss$output$plot

plot(ow.moss$intermediate$dat)
```



####Carry out ANOVA on all taxons

```{r}
fit <- aov(data = ht.data, formula = height ~ year*taxon)
anova(fit)
print(model.tables(fit), digits = 3)
tuk <- TukeyHSD(fit)
#tuk
class(tuk)
plot(tuk)
tuk$`year:taxon`

tuk.df <- data.frame(tuk$year) %>%
  rownames_to_column(var = "year") %>% 
  separate(col = year, into = c("year2", "year1"), sep = "-") %>% 
  mutate(year.diff = as.numeric(year2) - as.numeric(year1)) %>% 
  arrange(p.adj) %>% 
  filter(p.adj<0.05)
#tuk.df  


```


```{r}
fit.aov <- aov(data = dplyr::filter(ht.data, taxon == "Fo"), formula = height ~ year)
anova(fit.aov)
```

```{r}
fit.aov <- aov(data = dplyr::filter(ht.data, taxon == "grm"), formula = height ~ year)
anova(fit.aov)
```

```{r}
fit.aov <- aov(data = dplyr::filter(ht.data, taxon == "Vm"), formula = height ~ year)
anova(fit.aov)
```



> It can be seen from table 1 that the mean height of sheep’s fescue increased significantly (p<0.005) from 3.11 cm in 2003 to 3.78 cm in 2005 (the 2005 mean value has narrower associated confidence limits for pairwise comparisons with 2003 than the 2008 and 2010 values).  Further increase to 3.88 cm in 2008 and 3.89 in 2010 are however not significant, and the mean height of 2.86 cm in 2013 is significantly lower than that recorded in the three previous surveys, and not significantly different from the 2003 mean height.  The mean height of stiff sedge increased significantly (p<0.001) to 7.13 cm in 2005, from 4.2 cm in 2003.  In 2008 the main height was slightly lower than 2005 at 6.56 cm, before increasing to the highest recorded mean of 7.56 cm in 2010 and dropping slightly again to 6.94 cm in 2013, but these changes since 2005 are not significant. These mean height trends are also shown in figure 3.

> The mean height of all graminoid species (grasses and sedges) together are shown in figure 3 and table 1.  This shows a significant (p<0.001) increase from 3.28 cm in 2003 to 6.45 cm in 2010.  The mean on 5.75 cm in 2013 is slightly lower, but not significantly different.


 

> Figure 3. Mean height of key vegetation components: bilberry, sheep’s fescue, stiff sedge and moss and lichens


