---
title: 'Statistics Case Study: Cross Fell Grazing 2013: Data Preparation'
author: "CK"
date: "7 Sept 2017"
output: html_notebook
---

###Time run
Time run: `r Sys.time()`

###Packages
```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
```

##Data preparation

###Import data - cover part
Import the csv file provided
```{r, echo=T}
raw <- read.csv(file = "CROSS FELL COVER ALL.csv", sep = ",", quote = "",  header = T, skipNul = T)
```


###Wrangle data - cover
####Names
lets have a look at the names: 
```{r}
names(raw)
```
They are a bit of a mess.  

We can strip out a lot automatically, and replace some of the more exotic characters. Here I will remove the periods and replace `VM13.` with `Vm13` as names in **R** are case sensitive.   
```{r}
names(raw) <- gsub("\\.", "", names(raw))
colnames(raw)[5] <- c("Vm13")
#colnames(raw)
```

####Strip
We've got some unnecessary summary columns: They all start with `C`.  We'll strip them out: 
```{r}
raw[,grep("C", names(raw))] <- NULL
#names(raw)
```

DM has advised:

> Vmb – b indicates a version of the data with a couple of outliers – 10% cover – removed in 2008

So I am going to use the version without outliers (Vmb) as this is what DM did.  I'll delete all the Vm columns, and I'll subsequently rename Vmb to Vm:

```{r}
raw[c("Vm03", "Vm05", "Vm08", "Vm10", "Vm13")] <- NULL
colnames(raw)[1:5] <- c("Vm03", "Vm05", "Vm08", "Vm10", "Vm13")
#names(raw)
```


Finally we've got the natural log of moss, which we'll remove, because it is a derrived value and at this stage we only want the observations. 
```{r}
raw[,grep("lnmoss", names(raw))] <- NULL
names(raw)
```



####Make tidy
Now lets put this into [**tidy**](http://r4ds.had.co.nz/tidy-data.html#tidy-data-1) format:
  
> 1. Each **variable**    must have its own **column**.  
> 2. Each **observation** must have its own **row**.  
> 3. Each **value**       must have its own **cell**.  
  
In this case the observations are of percent cover, and they key (currently column names) define a taxon and a year.  
```{r}
cov.data <- gather(data = raw, key = "tax.surv", value = "cover", na.rm = TRUE, factor_key = TRUE)
head(cov.data) #have a look
dim(cov.data) #get dimensions
levels(cov.data$tax.surv) #check all the variables are there
```

Now we need to split the `tax.surv` column. 
```{r}
cov.data <- separate(data = cov.data, col = tax.surv, into = c("taxon", "year"), sep = -3)
cov.data$taxon <- as.factor(cov.data$taxon)
levels(cov.data$taxon)
head(cov.data)
summary(cov.data)
```

Finally sort out the `year` column:
```{r}
cov.data$year[which(cov.data$year=="03")] <- 2003
cov.data$year[which(cov.data$year=="05")] <- 2005
cov.data$year[which(cov.data$year=="08")] <- 2008
cov.data$year[which(cov.data$year=="10")] <- 2010
cov.data$year[which(cov.data$year=="13")] <- 2013

cov.data$year <- as.factor(cov.data$year)
levels(cov.data$year) <- c("2003", "2005", "2008", "2010", "2013")
```

we good?
```{r}
summary(cov.data)
```
**we good.**
  
  
  
###Import data - height 
Import the csv file which I previously opened in Minitab and saved (probably wrongly in Unicode)
```{r, echo=T}
raw <- read.csv(file = "CROSS FELL HEIGHT ALL.csv", sep = ",", quote = "",  header = T, skipNul = T)
```

###Wrangle data - height
####Names
lets have a look at the names: 
```{r, echo=T}
names(raw)
```
They are a bit of a mess.  
We can strip out a lot automatically, and replace some of the more exotic characters.  Here I will remove the `X.` and the periods, and replace the first column name entirely.  
```{r}
names(raw) <- gsub("ht", "", names(raw))
names(raw) <- gsub("X.", "", names(raw))
names(raw) <- gsub("\\.", "", names(raw))
names(raw)
```

####Strip
We've got some unnecessary summary columns: They all start with `C`.  We'll strip them out: 
```{r}
raw[,grep("C", names(raw))] <- NULL
#names(raw)
```


####Make tidy
Now lets put this into **tidy** format.

In this case the observations are of percent cover, and they key (currently column names) define a taxon and a year.  
```{r}
ht.data <- gather(data = raw, key = "tax.surv", value = "height", na.rm = TRUE, factor_key = TRUE)
head(ht.data) #have a look
dim(ht.data) #get dimensions
levels(ht.data$tax.surv) #check all the variables are there
```

Now we need to split the `tax.surv` column. 
```{r}
ht.data <- separate(data = ht.data, col = tax.surv, into = c("taxon", "year"), sep = -3)
ht.data$taxon <- as.factor(ht.data$taxon)
levels(ht.data$taxon)
head(ht.data)
summary(ht.data)
```

Finally sort out the `year` column:
```{r}
ht.data$year[which(ht.data$year=="03")] <- 2003
ht.data$year[which(ht.data$year=="05")] <- 2005
ht.data$year[which(ht.data$year=="08")] <- 2008
ht.data$year[which(ht.data$year=="10")] <- 2010
ht.data$year[which(ht.data$year=="13")] <- 2013

ht.data$year <- as.factor(ht.data$year)
levels(ht.data$year) <- c("2003", "2005", "2008", "2010", "2013")
```

*we good?*
```{r}
summary(ht.data)
```
**we good.**
  
  
###Import data - all quadrat data

  *we can't do this at the moment because we don't have the quadrat data*

  ###Merge data
  

###Contextual data
  
  ####Quadrats
  
  Number of quadrats surveyed per year (taken from report)
```{r}
quad.n <- data.frame(year = c("2003", "2005", "2008", "2010", "2013"), n.total = c(42, 44, 20, 31, 56))
```

##Save and clean up

```{r}
#save to file
save(cov.data, ht.data, quad.n, file = "CrossFell.data.prepd.rda")
#cleanup environment
rm(cov.data, ht.data, quad.n, raw)
```

